\documentclass{article}

\usepackage[utf8]{inputenc}
\usepackage{parskip}
\usepackage{amssymb,amsfonts,amsmath,amscd,amsthm}
\usepackage{bm}
\usepackage{hyperref}
\usepackage[pdftex]{graphicx}
\usepackage{url}
\usepackage[usenames,dvipsnames]{color}
\usepackage{enumitem}
\usepackage{mathtools}
\usepackage{float}

% for typesetting in-text numbers and units
\usepackage{siunitx}

% improved typesetting
\usepackage{microtype}

% Useful commands.
\input{commands}

\title{Model Predictive Control}
\author{Adam Heins}

\begin{document}

\maketitle

Here we derive some basic model predictive controllers.

\section{Linear MPC}

Suppose we have a discrete-time linear time-invariant system of the form
\begin{equation*}
  \bm{x}_{k+1} = \bm{A}\bm{x}_k + \bm{B}\bm{u}_k,
\end{equation*}
where~$\bm{x}_k$ and~$\bm{u}_k$ are the state and input at the $k$th timestep,
respectively. Consider the optimal control problem
\begin{equation}\label{eq:ocp}
  \begin{aligned}
    \minimize &\quad \ell_f(\bm{x}_N) + \sum_{k=0}^{N-1} \ell(\bm{x}_k,\bm{u}_k) \\
    \text{subject to} &\quad \bm{x}_{k+1} = \bm{A}\bm{x}_k + \bm{B}\bm{u}_k, \quad k=0,\dots,N, \\
                      &\quad \bm{u}_k\in\mathcal{U}, \quad k=0,\dots,N-1, \\
                      &\quad \bm{x}_k\in\mathcal{X}, \quad k=0,\dots,N-1, \\
                      &\quad \bm{x}_N\in\mathcal{X}_f,
  \end{aligned}
\end{equation}
where we are trying to find the control inputs~$\bm{u}$ that minimize the total
cost (running cost~$\ell$ and terminal cost~$\ell_f$) over a horizon of~$N$
steps subject to the dynamics and state and input constraints.

Let us consider how to solve~\eqref{eq:ocp}. Define the \emph{lifted} state and
input vectors
\begin{align*}
  \bar{\bm{x}} &= \begin{bmatrix} \bm{x}_0^T,\dots,\bm{x}_N^T \end{bmatrix}^T, &
  \bar{\bm{u}} &= \begin{bmatrix} \bm{u}_0^T,\dots,\bm{u}_{N-1}^T \end{bmatrix}^T,
\end{align*}
and let us take the costs to be
\begin{align*}
  \ell(\bm{x},\bm{u}) &= \bm{x}^T\bm{Q}\bm{x} + \bm{u}^T\bm{R}\bm{u}, \\
  \ell_f(\bm{x}) &= \bm{x}^T\bm{Q}\bm{x},
\end{align*}
for positive definite~$\bm{Q}$ and~$\bm{R}$. We have assumed that the terminal
cost is the same as the running cost for simplicity. We will take the input
constraints to be the bounds~$\mathcal{U}=\{\bm{u}\mid
\bm{u}_{\min}\leq\bm{u}\leq\bm{u}_{\max}\}$ and we will omit the state
constraints entirely. So far we have assumed that we are regulating the system
to the origin~$\bm{x}=\bm{0}$, but we will also discuss tracking a desired
trajectory~$\bar{\bm{x}}_d$.

\subsection{Multiple shooting}

In a multiple shooting approach, we optimize over both~$\bar{\bm{x}}$
and~$\bar{\bm{u}}$. Let us define the matrices
\begin{align*}
  \bar{\bm{A}} &= \begin{bmatrix} \bm{A} & -\bm{I} & \bm{0} \\ \bm{0} & \bm{A} & -\bm{I} \\ & & \ddots & \ddots \end{bmatrix} &
  \bar{\bm{B}} &= \diag(\bm{B})^{N} \\
  \bar{\bm{Q}} &= \diag(\bm{Q})^{N+1}, & \bar{\bm{R}} &= \diag(\bm{R})^N,
\end{align*}
where~$\bm{I}$ is the identity matrix and~$\diag(\bm{X})^n$ is a block diagonal
matrix with~$n$ blocks of the matrix~$\bm{X}$.
Given the actual value of the current state~$\hat{\bm{x}}_0$, the
OCP~\eqref{eq:ocp} can be formulated as the quadratic program
\begin{equation}\label{eq:mpc_ms}
  \begin{aligned}
    \minimize_{\bar{\bm{x}},\bar{\bm{u}}} &\quad \bar{\bm{x}}^T\bar{\bm{Q}}\bar{\bm{x}} + \bar{\bm{u}}^T\bar{\bm{R}}\bar{\bm{u}} \\
    \text{subject to} &\quad \bm{x}_0 = \hat{\bm{x}}_0 \\
                      &\quad \bar{\bm{A}}\bar{\bm{x}} + \bar{\bm{B}}\bar{\bm{u}} = \bm{0} \\
                      &\quad \bar{\bm{u}}_{\min} \leq \bar{\bm{u}} \leq \bar{\bm{u}}_{\max}.
  \end{aligned}
\end{equation}
If we want to track a desired trajectory~$\bar{\bm{x}}_d$, then we add the
linear term~$-2\bar{\bm{x}}_d^T\bar{\bm{Q}}\bar{\bm{x}}$ to the cost.

\subsection{Single shooting}

In a single shooting approach, we eliminate the state variables and optimize
only over~$\bar{\bm{u}}$. Let us (re)define the matrices
\begin{align*}
  \bar{\bm{A}} &= \begin{bmatrix} \bm{I} \\ \bm{A} \\ \bm{A}^2 \\ \vdots \\ \bm{A}^{N} \end{bmatrix}, &
  \bar{\bm{B}} &= \begin{bmatrix}
    \bm{0} & \bm{0} & \bm{0} & \hdots \\
    \bm{B} & \bm{0} & \bm{0} & \hdots \\
    \bm{A}\bm{B} & \bm{B} & \bm{0} \\
    \bm{A}^{2}\bm{B} & \bm{A}\bm{B} & \bm{B} \\
    \vdots & & & \ddots \\ \bm{A}^{N-1}\bm{B}
  \end{bmatrix},
\end{align*}
such that~$\bar{\bm{x}}=\bar{\bm{A}}\hat{\bm{x}}_0 + \bar{\bm{B}}\bar{\bm{u}}$.
The objective function becomes
\begin{equation*}
  \begin{aligned}
    \bar{\bm{x}}^T\bar{\bm{Q}}\bar{\bm{x}} + \bar{\bm{u}}^T\bar{\bm{R}}\bar{\bm{u}} &= (\bar{\bm{A}}\hat{\bm{x}}_0 + \bar{\bm{B}}\bar{\bm{u}})^T\bar{\bm{Q}}(\bar{\bm{A}}\hat{\bm{x}}_0 + \bar{\bm{B}}\bar{\bm{u}}) + \bar{\bm{u}}^T\bar{\bm{R}}\bar{\bm{u}} \\
                                                                                    &= \hat{\bm{x}}_0^T\bar{\bm{A}}^T\bar{\bm{Q}}\bar{\bm{A}}\hat{\bm{x}}_0 + 2\hat{\bm{x}}_0^T\bar{\bm{A}}^T\bar{\bm{Q}}\bar{\bm{B}}\bar{\bm{u}} + \bar{\bm{u}}^T(\bar{\bm{R}}+\bar{\bm{B}}^T\bar{\bm{Q}}\bar{\bm{B}})\bar{\bm{u}},
  \end{aligned}
\end{equation*}
where the first term is constant and so can be omitted from the optimization
problem. We have also eliminated the dynamics equality constraints,
so~\eqref{eq:ocp} just becomes the quadratic program
\begin{equation}\label{eq:mpc_ss}
  \begin{aligned}
    \minimize_{\bar{\bm{u}}} &\quad (1/2)\bar{\bm{u}}^T(\bar{\bm{R}}+\bar{\bm{B}}^T\bar{\bm{Q}}\bar{\bm{B}})\bar{\bm{u}} + \hat{\bm{x}}_0^T\bar{\bm{A}}^T\bar{\bm{Q}}\bar{\bm{B}}\bar{\bm{u}} \\
    \text{subject to} &\quad \bar{\bm{u}}_{\min} \leq \bar{\bm{u}} \leq \bar{\bm{u}}_{\max}.
  \end{aligned}
\end{equation}
If we want to track a desired trajectory~$\bar{\bm{x}}_d$, then we modify the
linear term of the cost
to~$(\bar{\bm{A}}\hat{\bm{x}}_0-\bar{\bm{x}}_d)^T\bar{\bm{Q}}\bar{\bm{B}}\bar{\bm{u}}$.

\section{Nonlinear MPC}

Suppose now we have a discrete-time nonlinear system of the form
\begin{equation}\label{eq:nonlinear}
  \bm{x}_{k+1} = \bm{f}(\bm{x}_k,\bm{u}_k).
\end{equation}
We will proceed in a similar way as the linear MPC case by \emph{linearizing}
the system about a trajectory~$(\tilde{\bm{x}},\tilde{\bm{u}})$, yielding
\begin{align*}
  \bm{A}_k &= \pd{\bm{f}}{\bm{x}}\evalat_{(\tilde{\bm{x}}_k,\tilde{\bm{u}}_k)}, &
  \bm{B}_k &= \pd{\bm{f}}{\bm{u}}\evalat_{(\tilde{\bm{x}}_k,\tilde{\bm{u}}_k)}.
\end{align*}
At each timestep, our approach will be to iteratively linearize about some
nominal trajectory, then formulate and solve a QP similar to the above.
However, differently from the linear case, we will iteratively re-linearize and
solve the QP, until the solution has converged (or we have run out of compute
time). That means that each QP will solve for an update to the current
solution, rather than the solution directly. We will use a multiple shooting
approach.

Let~$\tilde{\bm{x}}$ and~$\tilde{\bm{u}}$ be the lifted state and input
trajectories from the previous iteration (we can initialize~$\tilde{\bm{u}}$
from the previous timestep or simply with all zeros, then roll
out~$\tilde{\bm{x}}$ using the nonlinear dynamics~\eqref{eq:nonlinear}). Then
we want to solve for updates~$\delta\bar{\bm{x}}$ and~$\delta\bar{\bm{u}}$ to
obtain a new optimal trajectory
\begin{align*}
  \bar{\bm{x}} &= \tilde{\bm{x}} + \delta\bar{\bm{x}}, \\
  \bar{\bm{u}} &= \tilde{\bm{u}} + \delta\bar{\bm{u}}.
\end{align*}
Let us define the matrices
\begin{align*}
  \bar{\bm{A}} &= \begin{bmatrix} \bm{A}_0 & -\bm{I} & \bm{0} \\ \bm{0} & \bm{A}_1 & -\bm{I} \\ & & \ddots & \ddots \\ & & & \bm{A}_{N-1} & -\bm{I} \end{bmatrix}, &
  \bar{\bm{B}} &= \begin{bmatrix} \bm{B}_0 & \bm{0} \\ \bm{0} & \bm{B}_1 \\ & & \ddots \\ & & & \bm{B}_{N-1} \end{bmatrix},
\end{align*}
Substituting into~\eqref{eq:mpc_ms}, we get
\begin{equation}\label{eq:mpc_nl}
  \begin{aligned}
    \minimize_{\delta\bar{\bm{x}},\delta\bar{\bm{u}}} &\quad (1/2)\delta\bar{\bm{x}}^T\bar{\bm{Q}}\delta\bar{\bm{x}} + (1/2)\delta\bar{\bm{u}}^T\bar{\bm{R}}\delta\bar{\bm{u}} + \tilde{\bm{x}}^T\bar{\bm{Q}}\delta\bar{\bm{x}} + \tilde{\bm{u}}^T\bar{\bm{R}}\delta\bar{\bm{u}}\\
    \text{subject to} &\quad \delta\bm{x}_0 = \bm{0} \\
                      &\quad \bar{\bm{A}}\delta\bar{\bm{x}} + \bar{\bm{B}}\delta\bar{\bm{u}} + \bar{\bm{A}}\tilde{\bm{x}} + \bar{\bm{B}}\tilde{\bm{u}}= \bm{0} \\
                      &\quad \bar{\bm{u}}_{\min}-\tilde{\bm{u}} \leq \delta\bar{\bm{u}} \leq \bar{\bm{u}}_{\max}-\tilde{\bm{u}},
  \end{aligned}
\end{equation}
which we can iterate at each timestep as desired.
If we want to track a desired trajectory~$\bar{\bm{x}}_d$, then we add the
linear term~$-\bar{\bm{x}}_d^T\bar{\bm{Q}}\delta\bar{\bm{x}}$ to the cost. Note
that if more complex constraints are included in the OCP, then we likely need
to employ a filter or line-search method to determine the optimal step size,
rather than taking a full step at each iteration.

\end{document}
